Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-06-13 draft/sections/soundness_and_completeness.tex
arxiv version added, repo restructured

@@

\begin{theorem}
$[-]$ is full.
\begin{proof}
As we quotient by $\mathcal{S}$ instead of arbitrary cospans we can only consider their normal forms $A_1 + \ldots A_n$ where all of $A_i$ do not contain any hierarchical edges.
Because each of $A_i$ is a plain hypergraph, they have a pre-image in $\textsf{PROP}(\Sigma)$, i.e. $A_i = \llbracket \cdot \rrbracket;\mathcal{I}(a_i)$.
And then $[\mathcal{J}(a_1) + \ldots + \mathcal{J}(a_n)] = [\mathcal{J}(a_1)] + \ldots + [\mathcal{J}(a_n)] = A_1 + \ldots + A_n$ and hence $\mathcal{J}(a_1) + \ldots + \mathcal{J}(a_n)$ is the pre-image of $A_1 + \ldots + A_n$.
\end{proof}
\end{theorem}

\question{Is proof ok?}

@@

\begin{theorem}
    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ in $\textsf{PROP}^{+}$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$.
\end{theorem}

>>>>

\begin{proposition}
    For each cospan $A$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$ there is a week normal form such that $A =_{\mathcal{S}} A_1 + \ldots A_m$ where each of $A_i$ does not contain any hierarchical edges.
    \begin{proof}
    By applying the rules 1-6 in $\mathcal{S}$ (the ones from Figure~\ref{fig:string-equations}) we either remove 1 hierarchical edge (rules 5-6) or do not change the number of them, but expand the edge onto other edges.
    We will have a redex for such a rule until there is only one hierarchical edge at the outermost level (and only one outermost edge) with no other hierarchical edges within it.
    \end{proof}
\end{proposition}

@@

\begin{theorem}
$[-]$ is full.
\begin{proof}
As we quotient by $\mathcal{S}$ instead of arbitrary cospans we can only consider their normal forms $A_1 + \ldots A_n$ where all of $A_i$ do not contain any hierarchical edges as per the proposition above.
Because each of $A_i$ is a plain hypergraph, they have a pre-image in $\textsf{PROP}(\Sigma)$, i.e. for all $i$ there exists $a_i$ in $\textsf{PROP}(\Sigma)$ such that $A_i = \llbracket a_i \rrbracket;\mathcal{I}$.
And then $[\mathcal{J}(a_1) + \ldots + \mathcal{J}(a_n)] = [\mathcal{J}(a_1)] + \ldots + [\mathcal{J}(a_n)] = A_1 + \ldots + A_n$ and hence $\mathcal{J}(a_1) + \ldots + \mathcal{J}(a_n)$ is the pre-image of $A_1 + \ldots + A_n$.
\end{proof}
\end{theorem}

@@

\begin{theorem}[Theorem 35~\cite{Frobenius2}]

    $a \Rightarrow_{R} b$ iff $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$.
    Where $\llangle \cdot \rrangle : \textsf{PROP}(\Sigma) + \textsf{Frob} \to \catname{Csp_{D}(Hyp_{\Sigma})}$ and $\Rrightarrow$ is a convex rewriting relation for morphisms in $\catname{Hyp_{\Sigma}}$.
\end{theorem}

\begin{remark}
    If $\llbracket a \rrbracket = n \xrightarrow{} A \xleftarrow{} m$ then $\llangle ^{\ulcorner} A ^{\urcorner} \rrangle$ is isomorphic to $0 \xrightarrow{} A \xleftarrow{} n + m$.
\end{remark}

\begin{proposition}
    $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$ iff $[\mathcal{J}(a)] \Rrightarrow_{[R]}^{+} [\mathcal{J}(b)]$.
     Note that $a$ and $b$ are from $\textsf{PROP}(\Sigma)$.
    \begin{proof}
        The forward direction follows by the remark above and by recalling that $\Rrightarrow$ implies there exists a pushout square below
        \[
        \scalebox{1}{
            \tikzfig{soundness_and_completeness/pushout_in_Hyp}
        }    
        \]
        where the matching $m$ is convex, $i+j \to C \to G$ is a boundary complement~\cite{Frobenius2} and marked squares are pushouts
        and by using the injection $\mathcal{K} : \catname{Hyp_{\Sigma}} \to \catname{EHyp_{\Sigma}}$ there exists the following pushout square
        \[
          \scalebox{0.75}{
            \tikzfig{soundness_and_completeness/pushout_in_EHyp}
          }  
        \]
        where $m$ is convex and down-closed (as there are no hierarchical edges) and $\mathcal{K}(i+j) \to C \to G$ is a boundary complement.
        This by definition means that $\mathcal{K}(n) \xrightarrow{} \mathcal{K}(n) \xrightarrow{} G \xleftarrow{} \mathcal{K}(m) \xleftarrow{} \mathcal{K}(m) = [\mathcal{J}(a)] \Rrightarrow_{[R]}^{+} = [\mathcal{J}(b)] = \mathcal{K}(n) \xrightarrow{} \mathcal{K}(n) \xrightarrow{} G \xleftarrow{} \mathcal{K}(m) \xleftarrow{} \mathcal{K}(m)$.
        The other direction follows by applying the same argument backwards.
    \end{proof}
\end{proposition}

\begin{theorem}
    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ in $\textsf{PROP}^{+}$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$.
    \begin{proof}
        $a \Rightarrow_{\langle l, r \rangle}^{+} b$ implies that $a$ is decomposable as $a_1 + \ldots + a' + \ldots + a_k$ and $b$ is decomposable as $a_1 + \ldots + b' + \ldots + a_k$ and $a' \Rightarrow_{\langle l, r \rangle} b'$.
        The latter implies by the Theorem 35~\cite{Frobenius2} that $\llangle \ulcorner a' \urcorner \rrangle \Rrightarrow_{\langle \llangle ^\ulcorner l ^\urcorner \rrangle, \llangle ^\ulcorner r ^\urcorner \rrangle \rangle} \llangle \ulcorner b' \urcorner \rrangle$ which means that $[a'] \Rrightarrow_{\langle [l], [r] \rangle}^{+} [b']$.
        This means that there exists a convex matching of $[l]$ within $[a']$ and hence there is a convex match of $[l]$ in $[a_1] + \ldots + [a'] + \ldots + [a_k]$ and that the pushout complement is $[a_1] + \ldots + C + \ldots + [a_k]$ where $C$ is the pushout complement for $[a'] \Rrightarrow_{\langle [l], [r] \rangle} [b']$.
        Similarly, the pushout is $[a_1] + \ldots [b'] + \ldots + [a_k] = [b]$.

        For the converse direction, $[a] \Rrightarrow_{\langle [l],[r] \rangle}^{+} [b]$ implies that $[a] = A_{1} + \ldots A' + \ldots + A_{n}$ and $[b] = A_{1} + \ldots B' + \ldots + A_{n}$ and $A' \Rrightarrow_{\langle [l],[r] \rangle}^{+} B'$ because $[l]$ and $[r]$ have a pre-image in $\textsf{PROP}(\Sigma)$ and because we quotient by $\mathcal{S}$.
        The latter means by the proposition above that $a' \Rightarrow_{\langle l, r \rangle} b'$ such that $A' = \llbracket a' \rrbracket; \mathcal{I}$ and $B' = \llbracket b' \rrbracket; \mathcal{I}$ and hence $a = a_1 + \ldots + \mathcal{I}(a') + \ldots + a_n \Rightarrow_{\langle l, r \rangle}^{+} a_1 + \ldots \mathcal{I}(b') + \ldots a_n = b$
    \end{proof}
    \update{It's again somewhat handwaving, do we need any supporing proofs for the above?}
\end{theorem}

### 2024-06-13 extended_abstract/EGraphsExtendedAbstract.tex
arxiv version added, repo restructured

\usepackage{amsmath}
\usepackage{stmaryrd}
% \usepackage{amssymb}
\usepackage{tikz}
\usepackage{tikz-network}
\usepackage{xspace}
\usepackage{xargs}
\usepackage[
  colorinlistoftodos,
  prependcaption,
  % disable,
]{todonotes}
\usepackage{fontenc}

%\usepackage{minted}

\usepackage{subcaption}
\usepackage{caption}

\usepackage{hyperref}

\usepackage{docmute}
\usepackage{hyperref}
\usepackage{proof}

@@

\usepackage{aliascnt}
% \newaliascnt{definition}{thm}
% \newaliascnt{remark}{thm}
% \newaliascnt{example}{thm}
\newtheorem{thm}{Theorem}
\newtheorem{remark}[thm]{Remark}

\theoremstyle{definition}
\newtheorem{definition}[thm]{Definition}

\newtheorem{example}[thm]{Example}
\newtheorem{proposition}[thm]{Proposition}

\input{macros}
\input{sample.tikzstyles}
\input{hypergraph.tikzstyles}
\input{hypergraph.tikzdefs}

@@

% \usepackage{mathtools,mathrsfs}
\usepackage{tikz-cd}
\usepackage{tikzit}
% to reduce tikz picture white space
\usepackage{adjustbox}

% for quotienting
% \usepackage{faktor}
\usepackage{xfrac}

\usepackage{import}

% category names
%\newcommand{\MdaCospans}{{\catname{MACsp_{D}(Hyp_{\Sigma})}}}

\newcommand\HypI[1]{\textbf{HypI}(#1)}
\newcommand\MdaCospans{\textbf{MHypI}(\Sigma)}
\newcommand{\Ecospans}{{\catname{EHypI(\Sigma)}}}
\newcommand{\MdaEcospans}{{\catname{MEHypI(\Sigma)}}}
\newcommand{\WellTypedMdaEcospans}{{\catname{MEHypI(\Sigma)}}}
% conflict
\newcommand{\hashtag}{{\#}}
\newcommand{\consistency}{{\smile}}
% There is already a remark in the preamble, but it does
% not work for some reason

@@

\usepackage{cleveref} % after hyperref!

@@

It will often be convenient to give presentations of SMTs in terms of string diagrams.
\begin{example}
\label{example:csmt}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
	\scalebox{0.8}{
  	 \tikzfig{categorical-semantics/Cartesian-equipment}
	}
\]
together with the following associativity, commutativity and unitality equations, again given in terms of string diagrams. 
\[
	\scalebox{0.6}{
	\tikzfig{categorical-semantics/Cartesian-theory}	
	}
\]
The Cartesian SMT is given by a set of generators $\Sigma_C$, together with the generators and equations of the SMT of commutative monoids, and the following additional naturality equations, for every $c \in \Sigma_C$
\[
	\scalebox{0.6}{
	\tikzfig{categorical-semantics/Cartesian-naturality}
	}
\]
By Fox's Theorem~\cite{fox},  $\textbf{SMT}(\Sigma_C, \mathcal{E}_C)$ is equivalent to the free Cartesian category over $\Sigma_C$. 
\end{example}

@@

\begin{figure}
    \[
        \scalebox{0.65}{
        \tikzfig{categorical-semantics/egraph-strings}
        }
    \]
    \captionsetup{belowskip=-5ex}
    \caption{String diagrams for semilattice enriched symmetric monoidal categories.}
    \label{fig:egraph-strings}
    \end{figure}

@@

\begin{figure*}
\[  
    \scalebox{0.75}{
	\tikzfig{categorical-semantics/egraph-strings-equations}
    }
\]
\caption{Equations for a  semilattice enriched symmetric monoidal category}
\label{fig:string-equations}
\end{figure*}

@@

\begin{figure}
    \[
        \scalebox{0.5}{
        \tikzfig{categorical-semantics/egraph-translation-2}
        }
    \]
    \caption{Example translation of acyclic e-graphs into string diagrams for semilattice enriched Cartesian categories. }
    \label{fig:e-graph-example}
\end{figure}

@@

We can then translate an arbitrary e-graph into an e-string diagram as shown below, noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{categorical-semantics/egraph-translation-1}
\]
Note, further, that in the Cartesian case we can restrict to generating morphisms $c$ to have a single output,  without loss of generality,  by applying the universal properties of the Cartesian product.  This means the (informal) translation given above is fully general for acyclic e-graphs.
We believe that e-graphs with cycles can be represented as terms in Traced Cartesian SMTs and accounting for a trace above seems to be routine.

@@

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.5}{
        \tikzfig{categorical-semantics/egraph-translation-step-by-step-a-b}
        }
    \]
    \caption{Example translation from $(a)$ to $(b)$.}
    \label{fig:e-graph-example-a-b}
\end{figure*}

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.55}{
        \tikzfig{categorical-semantics/e-graph-substitution}
        }
    \]
    \caption{E-graph explicit substitution example.}
    \label{fig:e-graph-substitution}
\end{figure*}

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.5}{
        \tikzfig{categorical-semantics/e-string-substitution-2}
        }
    \]
    \caption{E-string diagrammatic substitution example.}
    \label{fig:e-string-substitution}
\end{figure*}

\begin{figure*}[t!]
    \vspace{-3cm}
    \centering
    \[
        \hspace{1.3cm}
        \resizebox{0.8\textwidth}{!}{
        \tikzfig{categorical-semantics/egraph-translation-step-by-step-b-c}
        }
    \]
    \caption{Example translation from $(b)$ to $(c)$.}
    \label{fig:e-graph-example-b-c}
\end{figure*}

@@

\bibliographystyle{acm}
\bibliography{bibliography}

>>>>

\usepackage{stmaryrd}

@@

% \newtheorem{thm}{Theorem}
% \newtheorem{remark}[thm]{Remark}

% \theoremstyle{definition}
% \newtheorem{definition}[thm]{Definition}

% \newtheorem{example}[thm]{Example}
% \newtheorem{proposition}[thm]{Proposition}

\input{../macros}
\input{../preamble}
\input{../sample.tikzstyles}
\input{../hypergraph.tikzstyles}
\input{../hypergraph.tikzdefs}

@@

@@

@@

It will often be convenient to give presentations of SMTs in terms of string diagrams.
\begin{example}
\label{example:csmt}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
	\scalebox{0.8}{
  	 \tikzfig{../figures/categorical-semantics/Cartesian-equipment}
	}
\]
together with the following associativity, commutativity and unitality equations, again given in terms of string diagrams. 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-theory}	
	}
\]
The Cartesian SMT is given by a set of generators $\Sigma_C$, together with the generators and equations of the SMT of commutative monoids, and the following additional naturality equations, for every $c \in \Sigma_C$
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-naturality}
	}
\]
By Fox's Theorem~\cite{fox},  $\textbf{SMT}(\Sigma_C, \mathcal{E}_C)$ is equivalent to the free Cartesian category over $\Sigma_C$. 
\end{example}

@@

\begin{figure}
    \[
        \scalebox{0.65}{
        \tikzfig{../figures/categorical-semantics/egraph-strings}
        }
    \]
    \captionsetup{belowskip=-5ex}
    \caption{String diagrams for semilattice enriched symmetric monoidal categories.}
    \label{fig:egraph-strings}
    \end{figure}

@@

\begin{figure*}
\[  
    \scalebox{0.75}{
	\tikzfig{../figures/categorical-semantics/egraph-strings-equations}
    }
\]
\caption{Equations for a  semilattice enriched symmetric monoidal category}
\label{fig:string-equations}
\end{figure*}

@@

\begin{figure}
    \[
        \scalebox{0.5}{
        \tikzfig{../figures/categorical-semantics/egraph-translation-2}
        }
    \]
    \caption{Example translation of acyclic e-graphs into string diagrams for semilattice enriched Cartesian categories. }
    \label{fig:e-graph-example}
\end{figure}

@@

We can then translate an arbitrary e-graph into an e-string diagram as shown below, noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{../figures/categorical-semantics/egraph-translation-1}
\]
Note, further, that in the Cartesian case we can restrict to generating morphisms $c$ to have a single output,  without loss of generality,  by applying the universal properties of the Cartesian product.  This means the (informal) translation given above is fully general for acyclic e-graphs.
We believe that e-graphs with cycles can be represented as terms in Traced Cartesian SMTs and accounting for a trace above seems to be routine.

@@

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.5}{
        \tikzfig{../figures/categorical-semantics/egraph-translation-step-by-step-a-b}
        }
    \]
    \caption{Example translation from $(a)$ to $(b)$.}
    \label{fig:e-graph-example-a-b}
\end{figure*}

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.55}{
        \tikzfig{../figures/categorical-semantics/e-graph-substitution}
        }
    \]
    \caption{E-graph explicit substitution example.}
    \label{fig:e-graph-substitution}
\end{figure*}

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \scalebox{0.5}{
        \tikzfig{../figures/categorical-semantics/e-string-substitution-2}
        }
    \]
    \caption{E-string diagrammatic substitution example.}
    \label{fig:e-string-substitution}
\end{figure*}

\begin{figure*}[t!]
    \vspace{-3cm}
    \centering
    \[
        \hspace{1.3cm}
        \resizebox{0.8\textwidth}{!}{
        \tikzfig{../figures/categorical-semantics/egraph-translation-step-by-step-b-c}
        }
    \]
    \caption{Example translation from $(b)$ to $(c)$.}
    \label{fig:e-graph-example-b-c}
\end{figure*}

@@

\bibliographystyle{acm}
\bibliography{../bibliography}
